Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-05-01 products/manuscript/Manuscript.qmd
Part 6 Updates
Final proofreading of the manuscript

# Summary/Abstract
Our study investigates the association between unintended pregnancy rates and short-acting reversible contraception (SARC) use across low- and middle-income countries. Data from the Guttmacher Institute's Adding It Up project was used, which included information on reproductive-aged women (15-49) and various demographic factors. Single and multivariate linear regression models were employed to examine relationships between predictor variables and the outcome variable (unintended pregnancy rate).

The exploratory data analysis revealed a less skewed distribution of unintended pregnancies among SARC users compared to other contraceptive methods. This led researchers to focus on SARC use for the remainder of the study. The Latin America/Caribbean region displayed the strongest association with unintended pregnancy rates, followed by Asia. The Oceania and Europe region had weaker associations. Marriage status, both currently and never married, was inversely associated with unintended pregnancy rates. This means that as the percent of women who are currently married or never married increases, the unintended pregnancy rate decreases. Maternal mortality rates showed a positive correlation with the outcome.

A full model incorporating five predictors (total SARC users, region, marital status, and maternal mortality rates due to unintended pregnancy and abortion) was built. Despite performing the best among the evaluated models and being validated through cross-validation and the bootstrap method, the full model exhibited tendencies towards overfitting the data.

Lasso regression, a technique that performs variable selection, was employed to address these overfitting concerns. The analysis identified the total number of SARC users as the strongest predictor, followed by region. Marital status and maternal mortality rates displayed weaker negative associations with unintended pregnancy rates. While the direction of the association between abortion-related maternal mortality and unintended pregnancy rates flipped in the Lasso model, the coefficient (effect size) remained minimal.

@@

We will be looking for patterns in contraceptive methods and unintended pregnancies across all age groups (15-19, 20-24, 25-35, and 35-49) to investigate the effectiveness of three contraceptive methods: long acting reversible and sterilization,short acting reversible, and traditional. Financial, social, religious, and cultural factors are current challenges against efforts towards reducing unintended pregnancies [@jonathanmarcbearakCountryspecificEstimatesUnintended2022]. According to a cross-sectional survey done in 2017, contraceptive use was prevalent among women who had attained higher education, were in stable partnerships, and identified with the dominant ethnic group . Additionally, factors like previous pregnancies and immigration status did not appear to influence contraceptive use in this study [@rivetContraceptionUseAttitudes2021]. African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies. This could be due to the current legal status of contraception within Latin American countries. Free emergency contraception access exists, but regulations differ. Chile, Colombia, and Ecuador legally recognize access, while Nicaragua and Bolivia rely on Ministerial protocols. Argentina and Brazil lack legal recognition, but offer protocols and guides. Mexico requires provision to victims of sexual and domestic violence [@heviaLegalStatusEmergency2012]. Stable contraceptive use among high-risk women suggests reaching even lower national unintended pregnancy rates might be challenging [@kavanaughContraceptiveMethodUse2018].

## Description of data and data source
The data source for this project is the Guttmacher Institute, which is a well-known research organization focusing on improving sexual and reproductive health around the world. The data used here is from their Adding It Up project. The goals of this project include estimating the need for, impact of, and costs associated with providing sexual and reproductive health services. Low and middle income countries are the target audience for this research, specifically women of reproductive age (15-49). Some of the variables included in this dataset are modern contraceptive use, unintended pregnancies, unplanned births, and abortions averted. Observations are drawn from nationally representative surveys, including Demographic and Health Surveys, UNICCEF Multiple Indicator Cluster Surveys, US Centers for Disease Control Reproductive Health Surveys, Performance Monitoring for Action Surveys, and others. The UN Population Division is the source of population projections for women in this age group, and estimates of unintended pregnancies are from the Guttmacher Institute, WHO, and other authors, which have been adjusted to 2019. It is important to note the types of contraception included in each category defined by the Adding It Up project. Permanent methods include female and male sterilization, and long acting methods include IUDs and implants. These methods are combined for several variables and include a 'larcster' prefix or suffix to identify them throughout the dataset. Short-acting methods include hormonal pills, injectables, male and female condoms, emergency contraceptive pills, patches, rings, diaphragms, vaginal spermicide, and other supply methods, which is denoted with a 'sarc' prefix or suffix in the variable name. Traditional methods include periodic abstinence, withdrawal, abstinence, and breast-feeding, which is denoted with a 'trad' prefix or suffix in the variable name.

For the purpose of this project, we chose several numeric variables from the data source: percentages of pregnancy outcomes, rates of safe and unsafe abortions, percentage of various contraceptive usages, percentages of care received for various pregnancy complications, and costs for many variables including abortions and STIs. There are only three character variables that can be used to group the numeric variables: country name, region, and sub region.

@@

A full list of variables can be found in the processing-file in the code > processing-code folder.

We hypothesize that the percentage of unintended preggnancies among SARC users will vary between the five regions based on differing access to this type of contraception.

@@

# Methods 
Reproducing this project requires R, RStudio, and Microsoft word. The README files in each folder contain descriptions of the documents located in that folder to help guide users through the reproducible processes. Begin with the processing-file.qmd in the code > processing-code folder. This file loads the original dataset from the Guttmacher Institute's Adding It Up project, loads the codebook for the dataset, and glimpses at the data. We chose 30 variables to clean to find the percentage of unintended pregnancies based on each type of contraceptive methods: long acting reversible and sterilization methods, short acting reversible methods, traditional methods, and no contraceptive usage. After the missing data was explored, we chose to use the percentage of unintended pregnancies variables related to each method of birth control because the total variable had unexplained missingness. The processed data is saved in the data > processed-data folder as processeddata.rds.

The next step is exploratory data analysis. The eda.qmd file in the code > eda-code folder contains all the code for this process. We produced a couple summary tables before we began a simple exploration of trends in percentage of unintended pregnancies by country in our first set of histograms. The second set of histograms explore contraceptive usage methods by country. These graphs revealed outliers, so we standardized by population size for each country. The next sets of plots are line graphs showing total usage of contraceptive methods to percentage of unintended pregnancies associated with that method. The last two sets of plots contain line graphs of the relationship between the two variables stratified by region and violin plots displaying the relationship between percentage of unintended pregnancies and the categorical usage variable created during processing. All of these figures are displayed and summarized in the Supplementary-Material.qmd file in the products > manuscript > supplement folder. This process led us to focus on the percentage of unintended pregnancies among short acting reversible methods as our single outcome because this method is becoming increasingly popular, so we feel that results from this contraception method will be the most relevant. Additionally, the exploratory data analysis process showed that the distribution of this outcome variable is less skewed than the other options, and the stratification by region is clearer, which allows for better understanding. All exploratory figures are saved in the results > figures > exploratory-figures folder.

Step three is the statistical analysis and model fitting portion found in the statistical-analysis.qmd file in the code > analysis-code folder. We explore the single outcome `pct_upreg_sarc` based on 7 predictors with simple linear regression, total number of women using short acting reversible methods, region, percent of currently married women, percent of women who never married, total rate of maternal mortality, rate of maternal mortality by unintended pregnancy, and rate of maternal mortality by abortion. Result tables for each model and test are stored in the results > tables folder.

Step four is to critically analyze the performance of our models to determine which one performs the best in predicting our outcome, which is also done in the statistical-analysis.qmd file in the code > analysis-code folder. We chose 2 performance metrics to find for each of our 9 models: Root Mean Square Error (RMSE) and Mean Absolute Error (MAE). Once we decided on the final model based on RMSE and MAE, we use cross-validation to determine how well it will perform on data it has not seen before. This involves splitting the data into testing and training pieces , which allows us to compare the predictions created using both datasets. To examine the uncertainty in the model, we will use the bootstrapping method to resample our data 100 times to find confidence intervals for our predictions. Next, we used a LASSO regression to add a penalty to each parameter included in the model to determine if any variables should be removed to improve model performance. It is important to tune the hyperparameters associated with LASSO regression to determine which model fits best. We used a grid search to ensure that we try all combinations of the parameters in our grid. Lastly, we added a random forest model to attempt to improve model performance. We used another parameter tuning grid to allow for cross-validated tuning of 2 variables: `mtry`, the number of variable to possibly split at in each node, and `min_n`, the minimal node size. Each of these processes are crucial in determining which model performs best for our data. Result tables for each model and test are stored in the results > tables folder. Figures created from cross-validation, bootstrapping, LASSO regression, and random forest modeling are stored in the results > figure > final-figures folder.

The Manuscript.qmd file for this document can be found in the products > manuscript folder.

## Data aquisition
This research aims to analyze the factors influencing unintended pregnancies and contraception use in low-and-middle-income countries. The Guttmacher Institute's Adding It Up project provides the foundation, offering data on sexual and reproductive health services in these regions.Focusing on women of reproductive age (15-49), the study usese information from nationally representative surveys and population projections to estimate unintended pregnancy rates.

While data on the overall percentage of unintended pregnancies was unavailable, we strategically utilized data on rates related to specific birth control methods. This approach allows us to analyze the relationship between these rates and various social and health factors. Factors like the number of women utilizing different birth control methods, maternal death rates, and marital status will be meticulously examined. All 7 of these predictor variables were found within the study's codebook, which can can be found in the `Data` folder of this project. By discovering these relationships, the research aims to shed light on the complex landscape of contraception use and unintended pregnancy in low-and -middle-income countries. This knowledge can empower policymakers and healthcare providers to make informed decisions that ultimately improve reproductive health outcomes for women.

## Data import and cleaning
As shown in the processing-file.qmd in the code > processing-code folder, we import the data from a .csv file that was downloaded directly from the Guttmacher AIU project website. We also included the codebook in this file. After taking a glimpse at the data, we identified the variables we need to clean and created a new dataset with those variables only. We explored missingness in the data, which is further explained in the supplementary-material.qmd file found in the products > manuscript > supplement folder. Simple processing tasks include converting variable types, converting decimal value into true percentages, creating variables to find total usage of each birth control method across age groups, exploring outliers, and creating new standardized total usage values based on population size.

## Statistical analysis
We began with a simple linear regression model with each of the 7 predictor variables to determine the association between each variable and the outcome. We explored the region variable further because our main goal is to understand the difference in outcome by region. This led to discovering correlations between some of the predictors, which we examined through correlation coefficients. We tried out a combination of these correlated predictors in 2 different models to further examine the interactions. After considering which predictors are most impactful on our outcome, we narrowed the final predictors down to a list of 5. We included a multiple linear regression model with all 5 of these predictors. We used the RMSE and MAE performance metrics to evaluate all 9 models that we created to determine which one fit best. Lastly, we used cross-validation to determine how well our multiple linear regression model performs on data it has not seen before. We also used LASSO regression and random forest modeling as forms of subset selection by applying a regularization penalty and generating a tree structure, respectively. Lastly, we examine the AIC values associated with our two best performing models.

@@

After deciding on a single outcome variable, we created a summary table with that outcome and the seven predictors we will use to predict it, which can be seen in @tbl-1. A summary table was done to present data about the seven predictor variables chosen in a structured format. Summary tables enable us to compare between different categories or groups within the data set. This can reveal trends or relationships. From the summary table below, it would appear that the region with the highest rate of unintended pregnancies due to SARC usage is Latin America/Caribbean at 24%, followed by Europe at 23%. Africa and Asia have a mean of 13% and 14% respectively, while Oceania appears to have the lowest rate at 6%. It appears that a higher percentage of women are currently married in Asia (66%) and Africa (60%) compared to Latin America/Caribbean (54%) and Oceania (60%).The average percentage of never married women across the 5 regions is 31%. Latin America/Caribbean region has the highest percentage of never married women at 35% and Asia region has has the lowest percentage of never married women at 29%.
```{r}
#| message: false
#| label: tbl-1
#| tbl-cap: "Table 1: Summary of All Explored Variables"
#| echo: FALSE
predictortable<- readRDS(here("results", "tables", "predictortable.rds"))
predictortable
```

We created a multitude of exploratory plots to understand our data better. The first exploratory result shown in @fig-1 displays the distribution of the outcome variable: percent of unintended pregnancies due to short acting reversible methods. The `pct_upreg_sarc` variable is slightly skewed to the right with an average around 12%.
```{r}
#| message: false
#| label: fig-1
#| fig-cap: "A bar chart was made to showcase the frequencies of the number of countries and the percentage of unintended pregnancy rates using SARC methods. The data presented suggests that there are fewer countries with a higher number of unintended pregnancies using SARCs considering more than half of the countries experience less than 20% of unintended pregnancies when using SARCs."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png"))
```

The next set of variables we explored were the total counts of women using each birth control method. There were two large outliers that prevented the graphs from being useful. After exploring the outliers, we standardized using the population size of reproductive aged women (15-49 years). There is a positive association between the percentage of unintended pregnancies among short acting reversible method users and the total number of women using that method of birth control, which is to be expected. Our outcome variable is the number of unintended pregnancies among women who are using short-acting reversible contraception, so the number of unintended pregnancies that can be classified in this outcome is dependent on the number of women using this type of contraception. This percentage should increase as the number of women using short acting-reversible contraception increases because more women are at risk of experiencing an unintended pregnancy due to this contraception if they are using this type of contraception in the first place. This outcome variable does not speak to the effectiveness of short-acting reversible contraception, so it makes sense that our outcome would increase as more women put themselves at risk of experiencing the outcome through becoming classified as short-acting reversible contraception users. As explained in the supplementary material, the dataset includes the percentage of unintended pregnancies among long-acting reversible contraception and sterilization users, traditional birth control users, and users who do not use a method of birth control. Each of these percentages are positively associated with number of people in its respective classification group that it represents. The association for the contraception that we chose to focus on for this analysis, short acting reversible contraception, is displayed in @fig-2.
```{r}
#| message: false
#| label: fig-2
#| fig-cap: "A Scatterplot was contructed to see if there is a correlation between a standardized count of women using SARCs and the percentage of unintended pregnancies. In this case, a positive correlation is shown. It suggests that as the number of women using SARCs increases (x-axis), the percentage of unintended pregnancies increases (y-axis)"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png"))
```

We stratified these four graphs by region to examine how the relationship differs across each of the regions which is shown in @fig-3 . The graph for short acting reversible methods shows clearer differences between regions, and Latin America/the Caribbean leads in the percentage of unintended pregnancies in this category, but Africa shows the sharpest increase in percentage of unintended pregnancies as the total population increases.
```{r}
#| message: false
#| label: fig-3
#| fig-cap: "Scatter plots are used to see if there is a relationship between two sets of continuous data. In this scatter plot, the relationship between the standardized count of women using SARCs and the percentage of unintended pregnancies is being examined.There appears to be a positive correlation between the standardized count of women using SARCs and the percentage of unintended pregnancies. The regions with the highest percentage of unintended pregnacies are shown to be Latin America & the Caribbean, followed by Africa."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png"))
```

We also created violin plots using the categorical usage variables for each birth control method as shown in @fig-4. It appears that the distribution of short acting reversible method usage is skewed towards lower percentages of unintended pregnancies across all usage levels, suggesting that more countries have lower percentage of unintended pregnancies among women using this method. The high usage countries appear to have a bottleneck in percentage of unintended pregnancies. The high and low usage countries appear to be skewed towards higher levels of unintended pregnancies, but the medium usage countries appear to have lower percentage of unintended pregnancies. 
```{r}
#| message: false
#| label: fig-4
#| fig-cap: "Violin plots are used to visualize the distribution of data across different categories. In this case, it is being used to see the distribution of unintended pregnancy rates among countries that use short-acting reversible contraceptives (SARCs).The violin plot below shows that the distribution of unintended pregnancy rates is wider for countries with high SARC usage compared to counties with medium or low SARC usage. This means that there is more variation in unintended pregnancy rates in counties with high SARC usage."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png"))
```

Lastly, we explored the predictor variables that will be used in our models. The Africa region accounts for most of the observations with about 53 countries, followed by Asia that includes about 35 countries. Latin America/the Caribbean accounts for about 25 countries. Europe and Oceania account for the least number of observations. The distribution of women currently married is approximately normal with an average around 57% and no major outliers. The distribution of women never married is slightly skewed to the right with an average around 30%, but there are a few outliers greater than 60%. The distribution of the total maternal mortality rate is highly skewed to the right with a couple outliers. We would have expected the highest rates to belong to India and China because they have the most people, but it is Chad, Sierra Leone, and South Sudan that account for the highest rates of total maternal mortality, which each belong to the African region. The distribution of the maternal mortality rate from unintended pregnancies is highly skewed to the right with a couple outliers, which is to be expected based on the trend shown in the previous histogram. Chad and South Sudan also account for the highest rates of maternal mortality in this category. The distribution of maternal mortality rate from abortions is also highly skewed to the right. The outliers are accounted for by Gambia and Mauritania in this predictor, which both belong to the African region.

@@

```{r}
#| message: false
#| label: tbl-2
#| fig-cap: "Table 2: This simple linear regresion model shows a strong effect of the standardized population size using short-acting reversible contraception methods on the outcome of percentage of unintended pregnancies due to short acting reversible contraception. This strong positive relationship indicates that as more women use this form of contraception, more women are susceptible to unintended pregancies due to SARCs because they are now at risk of experiencing the outcome. The coefficient of 87.65998 is the largest single coefficient of any predictor that we studied, which increases our confidence in including it as a predictor in our final model."
#| echo: FALSE
readRDS(here("results","tables", "resulttable1.rds")) 
```

>>>>

# Summary/Abstract
Our study investigates the association between unintended pregnancy rates and short-acting reversible contraception (SARC) use across low- and middle-income countries. Data from the Guttmacher Institute's Adding It Up project was used, which included information on reproductive-aged women (15-49) and various demographic factors. Single and multivariate linear regression models were employed to examine relationships between predictor variables and the outcome variable, percentage of unintended pregnancies among SARC users.

The exploratory data analysis revealed a less skewed distribution of unintended pregnancies among SARC users compared to other contraceptive methods. This led the researchers to focus on SARC use for the remainder of the study. Exploratory graphs with the region predictor showed that the Latin America/Caribbean region displayed the strongest association with percentage of unintended pregnancies, followed by Asia. The Oceania and Europe region had weaker associations. Marriage status, both currently and never married, was inversely associated with percentage of unintended pregnancies. As the percent of women who are currently married or never married increases, the percentage of unintended pregnancies decreases. Maternal mortality rates showed a positive correlation with the outcome.

A full model incorporating five predictors, total SARC users, region, marital status, and maternal mortality rates due to unintended pregnancy and abortion, was chosen as the ideal model for this data. Despite performing the best among the evaluated models and being validated through cross-validation and the bootstrap method, the full model exhibited tendencies towards overfitting the data.

Lasso regression, a technique that performs variable selection, was employed to address these overfitting concerns. The analysis identified the total number of SARC users as the strongest predictor, followed by region. Marital status and maternal mortality rates displayed weaker negative associations with unintended pregnancy rates. While the direction of the association between abortion-related maternal mortality and unintended pregnancy rates flipped in the Lasso model, the coefficient that represents the effect size remained minimal.

@@

We will be looking for patterns in contraceptive methods and unintended pregnancies across all age groups (15-19, 20-24, 25-35, and 35-49) to investigate the effectiveness of three contraceptive methods: long acting reversible and sterilization,short acting reversible, and traditional. Financial, social, religious, and cultural factors are current challenges against efforts towards reducing unintended pregnancies [@jonathanmarcbearakCountryspecificEstimatesUnintended2022]. According to a cross-sectional survey done in 2017, contraceptive use was prevalent among women who had attained higher education, were in stable partnerships, and identified with the dominant ethnic group. Additionally, factors like previous pregnancies and immigration status did not appear to influence contraceptive use in this study [@rivetContraceptionUseAttitudes2021]. African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies. This could be due to the current legal status of contraception within Latin American countries. Free emergency contraception access exists in this region, but differing regulations limit access. Chile, Colombia, and Ecuador legally recognize access to emergency contraception, while Nicaragua and Bolivia rely on Ministerial protocols. Argentina and Brazil lack legal recognition, but offer protocols and guides. Mexico requires provision of emergency contraception to victims of sexual and domestic violence [@heviaLegalStatusEmergency2012]. Stable contraceptive use among high-risk women suggests reaching even lower national unintended pregnancy rates might be challenging [@kavanaughContraceptiveMethodUse2018].

## Description of data and data source
The data source for this project is the Guttmacher Institute, which is a well-known research organization focusing on improving sexual and reproductive health around the world. The data used here is from their Adding It Up project. The goals of this project include estimating the need for, impact of, and costs associated with providing sexual and reproductive health services. Low and middle income countries are the target audience for this research, specifically women of reproductive age (15-49). Some of the variables included in this dataset are modern contraceptive use, unintended pregnancies, unplanned births, and abortions averted. Observations are drawn from nationally representative surveys, including Demographic and Health Surveys, UNICEF Multiple Indicator Cluster Surveys, US Centers for Disease Control Reproductive Health Surveys, Performance Monitoring for Action Surveys, and others. The UN Population Division is the source of population projections for women in this age group, and estimates of unintended pregnancies are from the Guttmacher Institute, WHO, and other authors, which have been adjusted to 2019.

It is important to note the types of contraception included in each category defined by the Adding It Up project. Permanent methods include female and male sterilization, and long acting methods include intrauterine devices (IUDs) and implants. These methods are combined for several variables and include a 'larcster' prefix or suffix to identify them throughout the dataset. Short-acting methods include hormonal pills, injectables, male and female condoms, emergency contraceptive pills, patches, rings, diaphragms, vaginal spermicide, and other supply methods, which is denoted with a 'sarc' prefix or suffix in the variable name. Traditional methods include periodic abstinence, withdrawal, abstinence, and breast-feeding, which is denoted with a 'trad' prefix or suffix.

For the purpose of this project, we chose several numeric variables from the data source: percentages of pregnancy outcomes, rates of safe and unsafe abortions, and percentage of various contraceptive usages. There are only three character variables that can be used to group the numeric variables: country name, region, and sub region.

@@

A full list of variables can be found in the `processing-file.qmd` in the code > processing-code folder.

We hypothesize that the percentage of unintended pregnancies among SARC users will vary between the five regions based on differing access to this type of contraception.

@@

# Methods 
Reproducing this project requires R, RStudio, and Microsoft word. The README files in each folder contain descriptions of the documents located in that folder to help guide users through the reproducible processes. Begin with the processing-file.qmd in the code > processing-code folder. This file loads the original dataset from the Guttmacher Institute's Adding It Up project, loads the codebook for the dataset, and glimpses at the data. We chose 30 variables to clean to find the percentage of unintended pregnancies based on each type of contraceptive methods: long acting reversible and sterilization methods, short acting reversible methods, traditional methods, and no contraceptive usage. After the missing data was explored, we chose to use the percentage of unintended pregnancies variables related to each method of birth control because the total variable had unexplained missing data. The processed data is saved in the data > processed-data folder as `processeddata.rds`.

The next step is exploratory data analysis. The eda.qmd file in the code > eda-code folder contains all the code for this process. We produced a couple summary tables before we began a simple exploration of trends in percentage of unintended pregnancies by country in our first set of histograms. The second set of histograms explore contraceptive usage methods by country. These graphs revealed outliers, so we standardized by population size for each country. The next sets of plots are scatter plots showing total usage of contraceptive methods and percentage of unintended pregnancies associated with that method. The last two sets of plots contain scatter plots of the relationship between the two variables stratified by region and violin plots displaying the relationship between percentage of unintended pregnancies and the categorical usage variable created during processing. All of these figures are displayed and summarized in the `Supplementary-Material.qmd` file in the products > manuscript > supplement folder. This process led us to focus on the percentage of unintended pregnancies among short acting reversible methods users as our single outcome because this method covers the most commonly used types of contraception, so we feel that results from this contraception method will be the most relevant. Additionally, the exploratory data analysis process showed that the distribution of this outcome variable is less skewed than the other options, and the stratification by region is clearer, which allows for better understanding. All exploratory figures are saved in the results > figures > exploratory-figures folder.

Step three is the statistical analysis and model fitting portion found in the `statistical-analysis.qmd` file in the code > analysis-code folder. We explore the single outcome `pct_upreg_sarc` based on 7 predictors: total number of women using short acting reversible methods, region, percent of currently married women, percent of women who never married, total rate of maternal mortality, rate of maternal mortality by unintended pregnancy, and rate of maternal mortality by abortion. We utilize simple linear regression and multiple linear regression with various combinations of the 7 predictors. Result tables for each model and test are stored in the results > tables folder.

Step four is to critically analyze the performance of our models to determine which one performs the best in predicting our outcome, which is also done in the `statistical-analysis.qmd` file in the code > analysis-code folder. We chose 2 performance metrics to find for each of our 9 models: Root Mean Square Error (RMSE) and Mean Absolute Error (MAE). Once we decided on the final model based on RMSE and MAE, we use cross-validation to determine how well it will performs on data it has not seen before. This involves splitting the data into testing and training pieces, which allows for comparison of the predictions created using both datasets. To examine the uncertainty in the model, we used the bootstrapping method to resample our data 100 times to find confidence intervals for our predictions. Next, we used a LASSO regression to add a penalty to each parameter included in the model to determine if any variables should be removed to improve model performance. It is important to tune the hyperparameters associated with LASSO regression to determine which model fits best. We used a grid search to ensure that all combinations of the parameters in our grid are considered. Lastly, we added a random forest model to attempt to improve model performance further. We used another parameter tuning grid to allow for cross-validated tuning of 2 variables: `mtry`, the number of variable to possibly split at in each node, and `min_n`, the minimal node size. Each of these processes are crucial in determining which model performed best for our data. Result tables for each model and test are stored in the results > tables folder. Figures created from cross-validation, bootstrapping, LASSO regression, and random forest modeling are stored in the results > figure > final-figures folder.

The `Manuscript.qmd` file for this document can be found in the products > manuscript folder.

## Data aquisition
This research aims to analyze the factors influencing unintended pregnancies and contraception use in low-and-middle-income countries. The Guttmacher Institute's Adding It Up project provides the foundation, offering data on sexual and reproductive health services in these regions. Focusing on women of reproductive age (15-49), the study uses information from nationally representative surveys and population projections to estimate percentage of unintended pregnancies.

While data on the overall percentage of unintended pregnancies was unavailable, we strategically utilized data on percentages related to specific birth control methods. This approach allows us to analyze the relationship between these percentages and various social and health factors. For example, the number of women utilizing different birth control methods, maternal death rates, and marital status are all meticulously examined. All 7 of these predictor variables were found within the study's codebook, which can can be found in the `data` folder of this project. By discovering these relationships, the research aims to shed light on the complex landscape of contraception use and unintended pregnancy in low-and -middle-income countries. This knowledge can empower policymakers and healthcare providers to make informed decisions that ultimately improve reproductive health outcomes for women.

## Data import and cleaning
As shown in the `processing-file.qmd` in the code > processing-code folder, the data is imported from a .csv file that was downloaded directly from the Guttmacher AIU project website. The codebook is also included in this file. After taking a glimpse at the data, we identified the variables that need to be cleaned and created a new dataset with those variables only. We explored missingness in the data, which is further explained in the `supplementary-material.qmd` file found in the products > manuscript > supplement folder. Simple processing tasks include converting variable types, converting decimal value into true percentages, creating variables to find total usage of each birth control method across age groups, exploring outliers, and creating new standardized total usage values based on population size.

## Statistical analysis
We began with a simple linear regression model with each of the 7 predictor variables to determine the association between each variable and the outcome. We explored the `region` variable further because our main research question is to understand the difference in outcome by region. This led to discovering correlations between some of the predictors, which we examined through correlation coefficients. We tried out a combination of these correlated predictors in 2 different models to further examine the interactions. After considering which predictors are most impactful on our outcome, we narrowed the final predictors down to a list of 5. We included a multiple linear regression model with all 5 of these predictors. We used the RMSE and MAE performance metrics to evaluate all 10 models that we created to determine which one fit best, including the null model with no predictors to serve as a comparison. Lastly, we used cross-validation to determine how well our multiple linear regression model performs on data it has not seen before. We also used LASSO regression and random forest modeling as forms of subset selection by applying a regularization penalty and generating a tree structure, respectively. Lastly, we examine the AIC values associated with our two best performing models.

@@

After deciding on a single outcome variable, we created a summary table with that outcome and the seven predictors we used to predict it, which can be seen in @tbl-1. Based on the summary table below, it appears that the region with the highest rate of unintended pregnancies due to SARC usage is Latin America/Caribbean at 24%, followed by Europe at 23%. Africa and Asia have a mean of 13% and 14% respectively, while Oceania appears to have the lowest rate at 6%. It appears that a higher percentage of women are currently married in Asia (66%) and Africa (60%) compared to Latin America/Caribbean (54%) and Oceania (60%).The average percentage of never married women across the 5 regions is 31%. Latin America/Caribbean region has the highest percentage of never married women at 35%, and the Asia region has the lowest percentage of never married women at 29%. Rates of maternal death are much higher in Africa with an average total rate of maternal deaths due to unintended pregnancies of 1,656 per 100,000 live births. The Asia region experiences the second highest rate in this category with 701 maternal deaths due to unintended pregnancies per 100,000 live births. 
```{r}
#| message: false
#| label: tbl-1
#| tbl-cap: "Summary of All Explored Variables"
#| echo: FALSE
predictortable<- readRDS(here("results", "tables", "predictortable.rds"))
predictortable
```

We created a multitude of exploratory plots to understand our data better. The first exploratory result shown in @fig-1 displays the distribution of the outcome variable: percent of unintended pregnancies due to short acting reversible methods. The `pct_upreg_sarc` variable is slightly skewed to the right with an average around 12%.
```{r}
#| message: false
#| label: fig-1
#| fig-cap: "This bar chart showcases the frequencies of the number of countries and the percentage of unintended pregnancies using SARC methods. The data presented suggests that there are fewer countries with a higher number of unintended pregnancies using SARCs considering more than half of the countries experience less than 20% of unintended pregnancies when using SARCs."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png"))
```

The next set of variables we explored were the total counts of women using each birth control method. There were two large outliers that prevented the graphs from being useful. After exploring the outliers, we standardized using the population size of reproductive aged women (15-49 years). There is a positive association between the percentage of unintended pregnancies among short acting reversible method users and the total number of women using that method of birth control, which is to be expected. Our outcome variable is the number of unintended pregnancies among women who are using short-acting reversible contraception, so the number of unintended pregnancies that can be classified in this outcome is dependent on the number of women using this type of contraception. This percentage should increase as the number of women using SARCs increases because more women are at risk of experiencing an unintended pregnancy due to this contraception if they are using this type of contraception in the first place. This outcome variable does not speak to the effectiveness of short-acting reversible contraception, so it makes sense that our outcome would increase as more women put themselves at risk of experiencing the outcome through becoming classified as SARC users. As explained in the supplementary material, the dataset includes the percentage of unintended pregnancies among long-acting reversible contraception and sterilization users, traditional birth control users, and users who do not use a method of birth control. Each of these percentages are positively associated with the number of people in its respective classification group that it represents. The association for the contraception that we chose to focus on for this analysis, SARCs, is displayed in @fig-2.
```{r}
#| message: false
#| label: fig-2
#| fig-cap: "A scatter plot was contructed to examine if there is a correlation between the standardized count of women using SARCs and the percentage of unintended pregnancies among SARC users. In this case, a clear positive correlation is shown. It suggests that as the number of women using SARCs increases (x-axis), the percentage of unintended pregnancies increases (y-axis)"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png"))
```

The previous scatter plot is stratified by region to examine how the relationship differs across each of the regions which is shown in @fig-3 . The graph shows clear differences between regions, and Latin America/the Caribbean leads in the percentage of unintended pregnancies in this category, but Africa experiences the sharpest increase in percentage of unintended pregnancies as the total population increases.
```{r}
#| message: false
#| label: fig-3
#| fig-cap: "In this scatter plot, the relationship between the standardized count of women using SARCs and the percentage of unintended pregnancies is examined by region.There appears to be a positive correlation between the standardized count of women using SARCs and the percentage of unintended pregnancies regardless of region, but there is some variation between regions. The regions with the highest percentage of unintended pregnacies are Latin America & the Caribbean, followed by Africa."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png"))
```

Violin plots were created using the categorical SARC usage variable as shown in @fig-4. It appears that the distribution of SAR usage is skewed towards lower percentages of unintended pregnancies across all usage levels, suggesting that more countries have lower percentage of unintended pregnancies among women using this method. The high-usage countries appear to have a bottleneck in percentage of unintended pregnancies. The low-usage countries appear to be skewed towards lower levels of unintended pregnancies among SARC users, but the medium-usage countries appear to have slightly higher percentages of unintended pregnancies. 
```{r}
#| message: false
#| label: fig-4
#| fig-cap: "Violin plots are used to visualize the distribution of data across different categories. In this case, it is being used to see the distribution of unintended pregnancy rates among countries that use short-acting reversible contraceptives (SARCs). This violin plot shows that the distribution of percentage of unintended pregnancies is wider for countries with high SARC usage compared to counties with medium or low SARC usage. This means that there is more variation in unintended pregnancy rates in counties with high SARC usage."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png"))
```

Lastly, we explored the predictor variables that will be used in our models. The Africa region accounts for most of the observations with about 53 countries, followed by Asia that includes about 35 countries. Latin America/the Caribbean accounts for about 25 countries. Europe and Oceania account for the least number of observations. The distribution of women currently married is approximately normal with an average around 57% and no major outliers. The distribution of women never married is slightly skewed to the right with an average around 30%, but there are a few outliers greater than 60%. The distribution of the total maternal mortality rate is highly skewed to the right with a couple outliers. We would have expected the highest rates to belong to India and China because they have the most people, but it is Chad, Sierra Leone, and South Sudan that account for the highest rates of total maternal mortality, which each belong to the African region. The distribution of the maternal mortality rate from unintended pregnancies is highly skewed to the right, which is to be expected based on the trend shown in the previous histogram. Chad and South Sudan also account for the highest rates of maternal mortality in this category. The distribution of maternal mortality rate from abortions is also highly skewed to the right. The outliers are accounted for by Gambia and Mauritania in this predictor, which both belong to the African region.

@@

```{r}
#| message: false
#| label: tbl-2
#| fig-cap: "Table 2: This simple linear regresion model shows a strong effect of the standardized population size using short-acting reversible contraception methods on the outcome of percentage of unintended pregnancies due to SARCs. This strong positive relationship indicates that as more women use this form of contraception, more women are susceptible to unintended pregancies due to SARCs because they are now at risk of experiencing the outcome. The coefficient of 87.65998 is the largest single coefficient of any predictor studied in this analysis, which increases our confidence in including it as a predictor in our final model."
#| echo: FALSE
readRDS(here("results","tables", "resulttable1.rds")) 
```
